---
title: "Measles across the world"
format: 
  dashboard:
    orientation: columns
    theme: united
    nav-buttons:
      - icon: github
        href: https://github.com/benparker9/measles-dashboard
        aria-label: GitHub
server: shiny
execute:
  warning: false
logo: "./www/measles.jpg" 
logo-alt: "Measles virus."
---

```{r}
#| label: load-packages
#| context: setup
library(tidyverse)
library(gt)
library(httr)
library(tidyjson)
library(jsonlite)
library(rvest)
library(here)
library(broom)
library(ggridges)
library(scales)
library(leaflet)
library(gganimate)
library(scales)
library(gifski)
library(plotly)
library(sf)
library(rsconnect)

rsconnect::setAccountInfo(name='cp-stats-541',
			  token='5F5597B8A1B0E4981EE9BD1C3C33F692',
			  secret='XH6GdyWeJdbZIaY16ZZES8UdvjhyEqsFqycKcF1q')

cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')

# need to do some cleaning for the webscraping process
cases_month2 <- cases_month |>
  mutate(month_abb = recode_values(month, 
                        1 ~ "Jan",
                        2 ~ "Feb",
                        3 ~ "Mar",
                        4 ~ "Apr",
                        5 ~ "May",
                        6 ~ "Jun",
                        7 ~ "Jul",
                        8 ~ "Aug",
                        9 ~ "Sep",
                        10 ~ "Oct",
                        11 ~ "Nov",
                        12 ~ "Dec"),
      country_scrape = str_to_lower(country),
      country_scrape = str_replace_all(country_scrape, " ", "-"))

api_load <- read.csv("./data/country_demographic.csv")
# load csv climate data
scraped_data <- read.csv(here("data", "weather-scrape.csv")) 

cases_month3 <- left_join(cases_month2, scraped_data, by = c("month_abb" = "month",  "country_scrape" = "country" ), 
                          relationship = "many-to-many")
meta_data <- left_join(cases_month3,
                       api_load,
                       by = c("country" = "name"),
                       relationship = "many-to-many")

yemen <- meta_data |>
  filter(country == "Yemen") 

url <- "https://www.worldometers.info/gdp/yemen-gdp/"

result <- tryCatch({
  read_html(url)
}, error = function(e) {
  message("Failed to read GDP page: ", e$message)
  NULL
})

gdp_data <- result |>
  html_elements("table") |>
  html_table()

clean_me <- function(df) {
  df |>
    rename(GDP_USD = 2) |>
    select(Year, GDP_USD) |>
    mutate(
      GDP_USD = str_remove_all(GDP_USD, "\\$"),
      GDP_USD = str_remove_all(GDP_USD, ","),
      GDP_USD = as.numeric(GDP_USD)
    )
}

gdp_wb <- clean_me(as.data.frame(gdp_data[1]))

gdp_imf <- clean_me(as.data.frame(gdp_data[3])) |>
  filter(Year < 2014)

yemen_gdp <- bind_rows(gdp_wb, gdp_imf)

bomb <- read.csv("yemen-bombing.csv") 

bomb2 <- bomb |>
  mutate(
    Date = dmy(Date),
    month = month(Date),
    year = year(Date)
  ) |>
  group_by(month, year) |>
  summarize(
    bombing_events = n(),
    total_casualty = sum(Fatalities + Woman.fatalities, na.rm = TRUE),
    .groups = "drop"
  )

yemen_measles_yearly <- yemen |>
  group_by(year) |>
  summarize(
    measles_total = sum(measles_total, na.rm = TRUE),
    rubella_total = sum(rubella_total, na.rm = TRUE),
    .groups = "drop"
  )

yemen_yearly <- bomb2 |>
  group_by(year) |>
  summarize(
    bombing_events = sum(bombing_events, na.rm = TRUE),
    total_casualty = sum(total_casualty, na.rm = TRUE),
    .groups = "drop"
  ) |>
  right_join(yemen_measles_yearly, by = "year") |>
  left_join(yemen_gdp, by = c("year" = "Year")) |>
  arrange(year)

yemen_dis <- read_sf('yem_admin1.geojson') |> select(adm1_name, geometry) 

bomb_dis <- bomb |>
  group_by(Governorate) |>
  summarize(bombing_events = n()) |>
  mutate(Governorate = recode(Governorate,
    "Dhalie"    = "Ad Dali'",
    "Bayda"     = "Al Bayda",
    "Hudaydah"  = "Al Hodeidah",
    "Jawf"      = "Al Jawf",
    "Maharah"   = "Al Maharah",
    "Mahwit"    = "Al Mahwit",
    "Marib"     = "Ma'rib",
    "Marib "    = "Ma'rib",    # trailing space
    "Saada"     = "Sa'dah",
    "Sanaa"     = "Sana'a",
    "Capital"   = "Sana'a City",
    "Taiz"      = "Ta'iz",
    "lahj"      = "Lahj"
  ))

district <- yemen_dis |>
  left_join(bomb_dis, by = c("adm1_name" = "Governorate"))


```

```{css}
.question-card {
  background-color: #FFCC99;
}

.question-card .card-header {
  background-color: #FF9966;
  font-size: 1.25rem;
  font-weight: 600;
}

.question-text {
  font-size: 1.8rem;
  line-height: 1.35;
}
```

# World Analysis {orientation="rows" scrolling="true"}

```{r}

#| label: world-prep
#| include: false

richest_5 <- c("Singapore", "Ireland", "Luxembourg", "Norway", "Qatar")
poorest_5 <- c("Burundi", "Central African Republic", "South Sudan", "Yemen", "Mozambique")

timeseries_df <- meta_data |>
  mutate(
    date = make_date(year, month, 1),
  measles_per_100k = (measles_total / population) * 100000  ) |>
  filter(country %in% c(richest_5, poorest_5)) |>
  mutate(
    wealth_group = case_when(
      country %in% richest_5 ~ "Richest 5",
      country %in% poorest_5 ~ "Poorest 5"
    )
  )
```

## Row {.flow}

::: {.card .question-card title="Guiding Question" fill="false"}
<div class="question-text">
<strong>Which country-level factors are associated with higher measles rates?</strong>
</div>
:::

## Row {height="88%"}

### Column{width="55%"}

```{r}
#| fig-height: 8
#| fig-width: 15
#| title: "Worldwide Measles Cases in 2023"
#| fig-cap:  "This map shows measles cases per 100,000 people in 2023. Darker countries had higher population-adjusted measles rates. Countries are colored by measles cases per 100,000 people, making countries with different population sizes easier to compare."

library(leaflet)
library(sf)
library(tidyverse)
library(rnaturalearth)


measles_country <- cases_year |>
  filter(year == 2023) |>
  group_by(iso3, country) |>
  summarise(total_cases = sum(measles_total, na.rm = TRUE),
            total_population = mean(total_population, na.rm = TRUE),
            cases_per_100k = (total_cases / total_population) * 100000)


world <- ne_countries(scale = "medium", returnclass = "sf")

world_measles <- world |>
  left_join(measles_country, by = c("iso_a3" = "iso3"))

pal <- colorNumeric(
  palette = "YlOrRd",
  domain = world_measles$cases_per_100k,
  na.color = "#cccccc"
)


leaflet(world_measles) |>
  addProviderTiles(providers$CartoDB.Positron,
                   options = providerTileOptions(noWrap = TRUE)) |>
  setView(lng = 0, lat = 20, zoom = 2) |>
  setMaxBounds(lng1 = -180, lat1 = -85,
               lng2 = 180,  lat2 = 85) |>
  addPolygons(fillColor = ~pal(cases_per_100k),
              color = "black",
              weight = 0.35,
              opacity = 0.8,
              popup = ~paste0("<strong>", country, "</strong><br>",
                              "Cases: ", total_cases, "<br>",
                              "Population: ", total_population, "<br>",
                              "Cases per 100k: ", round(cases_per_100k, 3))) |>
  addLegend(pal = pal,
            values = ~cases_per_100k,
            title = "Cases per 100k",
            position = "bottomright")

```

### Column {width="45%"}


::: {.card title = "Mealses Rates and Economic Factors"}
This dashboard compares **measles incidence**, **GDP per capita**, and **population-adjusted case rates** across countries from 2012–2025.


```{r}
#| warning: false
library(scales)

meta_data |>
  group_by(country, year) |>
  summarise(annual_measles = sum(measles_total, na.rm=TRUE),
            gdp_per_capita = first(gdp_per_capita),
            region = first(region.x),
            .groups = "drop") |>
  group_by(country) |>
  summarise(mean_annual_measles = mean(annual_measles),
            gdp_per_capita = first(gdp_per_capita),
            region = first(region),
            .grpups = "drop") |>
  ggplot(aes(x = gdp_per_capita, y = mean_annual_measles)) +
  geom_hex(bins = 20) +
  scale_x_log10(labels = dollar_format()) +
  scale_y_log10(labels = comma_format()) +
  scale_fill_gradient(low = "blueviolet",
                      high = "goldenrod") +
  labs(
    title = "GDP per Capita vs Mean Annual Measles Cases by Country",
    x = "GDP per capita (log scale)",
    y = "Mean annual measles cases (log scale)",
    fill = "Number of Countries",
    alt = paste(
      "Hexbin plot of countries comparing GDP per capita and mean annual measles
      cases.",
      "Both axes are on log scale.",
      "Contries are concentrated along a negatively correlated line, with many 
      contries having low GDP and high measles cases, while many others have low
      measles cases and high GDP."
    )
  ) +
  theme_bw() 

```

:::

::: {.card title = "Country Rankings and GDP"}

::: {.panel-tabset}

#### Measles Cases

```{r}
country_measles <- meta_data |>
  group_by(country) |>
  summarize(avg_yearly_measles_cases = mean(measles_total, na.rm = TRUE),
            avg_population = mean(population, na.rm = TRUE)) |>
  mutate(cases_per_100k = (avg_yearly_measles_cases / avg_population) * 100000) |>
  filter(!is.na(cases_per_100k),
         is.finite(cases_per_100k),
         cases_per_100k != 0)

rank_type <- c("Highest", "Lowest")

get_rank <- function(rank_type) {
  if (rank_type == "Highest") {
    country_measles |>
      slice_max(cases_per_100k, n = 5, with_ties = FALSE) %>%
      mutate(group = "5 Highest Cases per 100,000")
  } else {
    country_measles |>
      slice_min(cases_per_100k, n = 5, with_ties = FALSE) %>%
      mutate(group = "5 Lowest Cases per 100,000")
  }
}

table_data <- map_dfr(rank_type, get_rank)

table_data |>
  gt(groupname_col = "group") |>
  cols_label(country = "Country",
             avg_yearly_measles_cases = "Avg. Yearly Measles Cases",
             avg_population = "Avg. Population",
             cases_per_100k = "Cases per 100,000 People") |>
  fmt_number(columns = c(avg_yearly_measles_cases,
                         avg_population,
                         cases_per_100k),
             decimals = 2) |>
  tab_header(title = "Countries with the Highest and Lowest Measles Cases per 100,000 People",
             subtitle = "Based on average yearly measles cases and average population") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_row_groups()) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_title()) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_source_note(
    source_note = "Countries are ranked by average yearly measles cases per 100,000 people. The table compares the five countries with the highest rates and the five countries with the lowest rates.")
```

#### GDP per Capita

```{r}
gdppc_for_countries <- function(data, rich_countries, poor_countries) {
  data |>
    mutate(
      gdp_per_capita = (gdp / population) * 1000,
      group = case_when(
        country %in% rich_countries ~ "Richest 5",
        country %in% poor_countries ~ "Poorest 5")) |>
    filter(country %in% c(rich_countries, poor_countries)) |>
    group_by(group, country) |>
    arrange(desc(year), .by_group = TRUE) |>
    summarize(
      gdp_per_capita_usd = first(gdp_per_capita),
      .groups = "drop") |>
    arrange(group, desc(gdp_per_capita_usd)) |>
    gt(groupname_col = "group") |>
    fmt_currency(
      columns = gdp_per_capita_usd,
      currency = "USD",
      decimals = 0) |>
    cols_label(
      country = "Country",
      gdp_per_capita_usd = "GDP per Capita (USD)") |>
    tab_header(
      title = md("**GDP per Capita of Selected Richest and Poorest Countries**"),
      subtitle = "Countries grouped using GDP per capita rankings") |>
    tab_source_note(
      source_note = "GDP per capita was calculated as GDP divided by population.
      GDP appears to be stored in millions of USD, and population appears to be stored in thousands.") |>
    tab_options(table.width = pct(65))}

gdppc_for_countries(meta_data, richest_5, poorest_5)

```
:::
:::




# Yemen {orientation="rows"} 

## Descriptions {height="25%"}
**Why Yemen?**    
Yemen is an **outlier** country with unusually high measles rates. In conjunction, their socio-economic history is complex, with synergistic conditions across access to healthcare, GDP and war. Over the last three years, they have ranked in the top 5 countries with the highest measles rates, while falling in the bottom 5 countries per GDP. Additionally, they have had a [civil war since 2014](https://www.britannica.com/event/Yemeni-Civil-War).

## Rest {.fill}

### Yemen Bubble Plot {width="60%"}



```{r}
#| context: server
output$caption <- renderUI({
  links <- c(
    "bombing_events" = "https://yemendataproject.org/data/",
    "total_casualty" = "https://yemendataproject.org/data/",
    "measles_total"  = "https://www.who.int/data/gho",
    "rubella_total"  = "https://www.who.int/data/gho",
    "GDP_USD"        = "https://www.worldometers.info/gdp/yemen-gdp/"
  )
  
  captions <- c(
    "bombing_events" = "Yemen Data Project",
    "total_casualty" = "Yemen Data Project",
    "measles_total"  = "WHO Global Health Observatory",
    "rubella_total"  = "WHO Global Health Observatory",
    "GDP_USD"        = "Worldometer"
  )
  
  y_col <- input$y_var
  
  tags$p(
    tags$em("Source: "),
    tags$a(captions[[y_col]], href = links[[y_col]], target = "_blank")
  )
})

output$plot <- renderPlotly({
  y_col <- input$y_var
  y_data <- yemen_yearly[[y_col]]
  
  axis_labels <- c(
    "bombing_events" = "Bombing Events",
    "total_casualty" = "Total Casualties",
    "measles_total"  = "Measles Cases",
    "rubella_total"  = "Rubella Cases",
    "GDP_USD"        = "GDP (USD)"
  )
  
  plot_ly(
    data = yemen_yearly,
    x = ~year,
    y = y_data,
    type = "scatter",
    mode = "lines+markers"
  ) %>%
    layout(
      xaxis = list(title = "Year"),
      yaxis = list(title = axis_labels[[y_col]]),
      hovermode = "x unified"
    )
})

pal <- colorNumeric("viridis", domain = district$bombing_events)

output$yemen_map <- renderLeaflet({
  leaflet(district) |>
    addProviderTiles("Esri.WorldStreetMap") |>
    setView(lng = 47.5, lat = 15.5, zoom = 5) |>
    addPolygons(fillColor = ~pal(bombing_events),
                color = "black",
                weight = 0.35,
                opacity = 0.8,
                popup = ~paste0("<strong>", adm1_name, "</strong><br>",
                              "Bombing Events: ", bombing_events)) |>
    addLegend(pal = pal,
            values = ~bombing_events,
            title = "Total Bombing Events",
            position = "bottomright")

})


```

```{r}
selectInput("y_var", "Select Y Variable:",
  choices = c(
    "Bombing Events"   = "bombing_events",
    "Total Casualties from Bombing Events" = "total_casualty",
    "Measles Total"    = "measles_total",
    "Rubella Total"    = "rubella_total",
    "GDP (USD)"        = "GDP_USD"
  )
)

plotlyOutput("plot")
htmlOutput("caption")
```


### Country Choice {width="40%"}


```{r}

leafletOutput("yemen_map") 


# run  this in terminal to deploy app 
#rsconnect::deployApp("./")

```



