---
title: "Measles across the world"
format: 
  dashboard:
    orientation: columns
    embed-resources: true
    theme: united
    nav-buttons:
      - icon: github
        href: https://github.com/benparker9/measles-dashboard
        aria-label: GitHub
execute:
  warning: false
logo: measles.jpg 
logo-alt: "Meales virus."
---

```{r}
#| label: load-packages
library(tidyverse)
library(gt)
library(httr)
library(tidyjson)
library(jsonlite)
library(rvest)
library(here)
library(broom)
library(ggridges)
library(scales)
library(leaflet)
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')

# need to do some cleaning for the webscraping process
cases_month2 <- cases_month |>
  mutate(month_abb = recode_values(month, 
                        1 ~ "Jan",
                        2 ~ "Feb",
                        3 ~ "Mar",
                        4 ~ "Apr",
                        5 ~ "May",
                        6 ~ "Jun",
                        7 ~ "Jul",
                        8 ~ "Aug",
                        9 ~ "Sep",
                        10 ~ "Oct",
                        11 ~ "Nov",
                        12 ~ "Dec"),
      country_scrape = str_to_lower(country),
      country_scrape = str_replace_all(country_scrape, " ", "-"))

api_load <- read.csv("./data/country_demographic.csv")
# load csv climate data
scraped_data <- read.csv(here("data", "weather-scrape.csv")) 

cases_month3 <- left_join(cases_month2, scraped_data, by = c("month_abb" = "month",  "country_scrape" = "country" ), 
                          relationship = "many-to-many")
meta_data <- left_join(cases_month3,
                       api_load,
                       by = c("country" = "name"),
                       relationship = "many-to-many")


```

# World Analysis

## Col {width="50%"}

### GDP Plots + Model {height="50%"}

<p><strong>PLots about measles and GDP and model </strong><br>
Description.
</p>

:::panel-tabset

## Plot 1

```{r}

#| fig-alt: Line plot of measles cases over time comparing five richest countries and five poorest countries from Wikipedia ranking. 


richest_5 <- c("Singapore", "Ireland", "Luxembourg", "Norway", "Qatar")
poorest_5 <- c("Burundi", "Central African Republic", "South Sudan", "Yemen", "Mozambique")

timeseries_df <- meta_data |>
  mutate(
    date = make_date(year, month, 1),
  measles_per_100k = (measles_total / (population * 1000)) * 100000  ) |>
  filter(country %in% c(richest_5, poorest_5)) |>
  mutate(
    wealth_group = case_when(
      country %in% richest_5 ~ "Richest 5",
      country %in% poorest_5 ~ "Poorest 5"
    )
  )
ggplot(timeseries_df, aes(x = date, y = measles_per_100k, group = country, color = wealth_group)) +
  geom_line(linewidth = 1, alpha = 0.85) +
  scale_color_manual(
    values = c("Richest 5" = "steelblue4", "Poorest 5" = "firebrick3")
  ) +
  labs(
    title = "Adjusted Measles Rates Over Time (2012-2025)",
    subtitle = "Top 5 richest vs bottom 5 poorest countries",
    x = "Time",
    y = "Measles Cases per 100,000 People",
    color = "Group") +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom")
```

## Plot 2

```{r}
#| warning: false
library(scales)

meta_data |>
  group_by(country, year) |>
  summarise(annual_measles = sum(measles_total, na.rm=TRUE),
            gdp_per_capita = first(gdp_per_capita),
            region = first(region.x),
            .groups = "drop") |>
  group_by(country) |>
  summarise(mean_annual_measles = mean(annual_measles),
            gdp_per_capita = first(gdp_per_capita),
            region = first(region),
            .grpups = "drop") |>
  ggplot(aes(x = gdp_per_capita, y = mean_annual_measles)) +
  geom_hex(bins = 20) +
  scale_x_log10(labels = dollar_format()) +
  scale_y_log10(labels = comma_format()) +
  scale_fill_gradient(low = "blueviolet",
                      high = "goldenrod") +
  labs(
    title = "GDP per Capita vs Mean Annual Measles Cases by Country",
    x = "GDP per capita (log scale)",
    y = "Mean annual measles cases (log scale)",
    fill = "Number of Countries",
    alt = paste(
      "Hexbin plot of countries comparing GDP per capita and mean annual measles
      cases.",
      "Both axes are on log scale.",
      "Contries are concentrated along a negatively correlated line, with many 
      contries having low GDP and high measles cases, while many others have low
      measles cases and high GDP."
    )
  ) +
  theme_bw() 

```

## Model

```{r}
#| warning: false

linearreg_df <- meta_data |>
  select(
    country,
    year,
    gdp,
    population,
    measles_total,
    mean_temp,
    urban_population,
    tourists) |>
  mutate(
    gdp_per_capita = gdp / population,
    measles_rate_100k = (measles_total / population) * 100000,
    log_gdp_per_capita = log(gdp_per_capita),
    log_measles_rate_100k = log(measles_rate_100k + 1),
    log_tourists = log(tourists))

  lm(
    log_measles_rate_100k ~ log_gdp_per_capita + mean_temp + log_tourists + urban_population, 
    data = linearreg_df) |>
  tidy() |>
  gt() |>
  fmt_number(
    columns = c(estimate, std.error, statistic, p.value),
    decimals = 3
  ) |>
  cols_label(
    term = md("**Predictor**"),
    estimate = md("**Estimate**"),
    std.error = md("**Std. Error**"),
    statistic = md("**t-value**"),
    p.value = md("**p-value**")
  ) |>
  tab_header(
    title = md("**Regression Model Results**"),
    subtitle = "Predicting logged measles case rate per 100,000"
  ) |>
  tab_source_note(
    source_note = "Linear regression using GDP per capita, mean temperature, number of tourists, and urban population as predictors.")
```

:::

### GDP Tables {height="50%"}

<p><strong>Tables about measles and GDP </strong><br>
Description.
</p>

:::panel-tabset

## Table 1

```{r}
country_measles <- meta_data |>
  group_by(country) |>
  summarize(avg_yearly_measles_cases = mean(measles_total, na.rm = TRUE),
            avg_population = mean(population, na.rm = TRUE)) |>
  mutate(cases_per_100k = (avg_yearly_measles_cases / avg_population) * 100000) |>
  filter(!is.na(cases_per_100k),
         is.finite(cases_per_100k),
         cases_per_100k != 0)

rank_type <- c("Highest", "Lowest")

get_rank <- function(rank_type) {
  if (rank_type == "Highest") {
    country_measles |>
      slice_max(cases_per_100k, n = 5, with_ties = FALSE) %>%
      mutate(group = "5 Highest Cases per 100,000")
  } else {
    country_measles |>
      slice_min(cases_per_100k, n = 5, with_ties = FALSE) %>%
      mutate(group = "5 Lowest Cases per 100,000")
  }
}

table_data <- map_dfr(rank_type, get_rank)

table_data |>
  gt(groupname_col = "group") |>
  cols_label(country = "Country",
             avg_yearly_measles_cases = "Avg. Yearly Measles Cases",
             avg_population = "Avg. Population",
             cases_per_100k = "Cases per 100,000 People") |>
  fmt_number(columns = c(avg_yearly_measles_cases,
                         avg_population,
                         cases_per_100k),
             decimals = 2) |>
  tab_header(title = "Countries with the Highest and Lowest Measles Cases per 100,000 People",
             subtitle = "Based on average yearly measles cases and average population") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_row_groups()) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_title()) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_source_note(
    source_note = "Countries are ranked by average yearly measles cases per 100,000 people. The table compares the five countries with the highest rates and the five countries with the lowest rates.")
```
## Table 2
```{r}
gdppc_for_countries <- function(data, rich_countries, poor_countries) {
  data |>
    mutate(
      gdp_per_capita = (gdp / population) * 1000,
      group = case_when(
        country %in% rich_countries ~ "Richest 5",
        country %in% poor_countries ~ "Poorest 5")) |>
    filter(country %in% c(rich_countries, poor_countries)) |>
    group_by(group, country) |>
    arrange(desc(year), .by_group = TRUE) |>
    summarize(
      gdp_per_capita_usd = first(gdp_per_capita),
      .groups = "drop") |>
    arrange(group, desc(gdp_per_capita_usd)) |>
    gt(groupname_col = "group") |>
    fmt_currency(
      columns = gdp_per_capita_usd,
      currency = "USD",
      decimals = 0) |>
    cols_label(
      country = "Country",
      gdp_per_capita_usd = "GDP per Capita (USD)") |>
    tab_header(
      title = md("**GDP per Capita of Selected Richest and Poorest Countries**"),
      subtitle = "Countries grouped using GDP per capita rankings") |>
    tab_source_note(
      source_note = "GDP per capita was calculated as GDP divided by population.
      GDP appears to be stored in millions of USD, and population appears to be stored in thousands.") |>
    tab_options(table.width = pct(65))}

gdppc_for_countries(meta_data, richest_5, poorest_5)

```
:::



## Other Col {width="50%"}

### Ridge Plot of Regional distributions of Measles Cases {height="50%"}

<p><strong>Plots showing how environmental factors affect Measles Cases </strong><br>
Plot 1 is Ridge and plot 2 is heatmap.
</p>

:::panel-tabset

## Plot 1

```{r}
#| fig-height: 8
#| fig-width: 15
#| title: "A plot showing the distributions of measles cases per year across world regions."
#| fig-cap: "Figure 2: Comparison of the distribution of measles cases per year across world regions."

regions_df <- cases_year |>
  mutate(region = recode_values(region,
                    "EMRO" ~ "Eastern Mediterranean",
                    "AFRO" ~ "Africa",
                    "EURO" ~ "Europe",
                    "SEARO" ~ "South-East Asia",
                    "AMRO" ~ "Americas",
                    "WPRO" ~ "Western Pacific"))

regions_df |>
  filter(measles_total <= 5000) |>
  ggplot(aes(x = measles_total, y = fct_reorder(region, measles_total, 
                                                .desc = FALSE), )) +
  geom_density_ridges(aes(fill = region),
                      alpha = 0.5, scale = 1,
                      show.legend = FALSE) +
  labs(x = "Total Measles Cases per Year", 
       y = "", 
       title = "Comparison of Measles Cases across World Regions"
       ) +
  theme(legend.position = "none") +
  scale_x_continuous(labels = label_comma()) +
  theme_bw() +
  scale_fill_viridis_d(option = "B")

```
## Plot 2
```{r}
#| fig-height: 8
#| fig-width: 15
#| title: "A plot showing the relationship between climate, natural resources and emissions data and measles/rubella cases."
#| fig-cap: "Figure 1: Correlation matrix showing the relationship between climate, natural resources and emissions data and measles/rubella cases."


library(corrr)
make_titles <- function(x){
  # expects a character or factor as an input
  stopifnot(is.character(x) | is.factor(x))
  
  # remove the units at the end of the name
  str_remove(x, pattern = "(g|mm)$") |> 
  # replace all _ with spaces
  str_replace_all(pattern = "_", replacement = " ") |> 
  # remove extra whitespace on left 
  str_trim(side = "both") |> 
  # convert the first letter of each word to upper case
  str_to_title()
}

meta_data |>
  select(month, measles_total, measles_epi_linked, rubella_total, 
         rubella_epi_linked, mean_temp, mean_precip, surface_area,
         forested_area, co2_emissions, threatened_species, internet_users) |>
  correlate(method = "pearson",
            use = "pairwise.complete.obs") |>
  rename(term1 = term) |> 
  pivot_longer(cols = -term1, 
               names_to = "term2", 
               values_to = "correlation") |>
  mutate(
    across(.cols = c(term1, term2), 
           .fns = ~ make_titles(.x)
           )
    ) |> 
  ggplot(mapping = aes(x = term1,
                       y = term2, 
                       fill = correlation)) +
  geom_tile() +
  labs(x = "", 
       y = "", 
       title = "Correlation Between Measles/Rubella Cases and Biological Factors", 
       fill = "Correlation"
       ) +
  theme_bw() +
  scale_fill_viridis_c(option = "B")
```

:::

### Yoo {height="50%"}

<p><strong>Top 15 countries with the highest total measles cases (2023-2025)</strong><br>
Countries sorted in descending order of total measles cases.
</p>
::: panel-tabset 

## 2023 

```{r}
#| label: twentythree 

cases_year |>
  filter(year == 2023) |>
  select(country, measles_total, rubella_total) |>
  arrange(desc(measles_total)) |>
  slice_max(measles_total, n=15) |>
  gt() |>
  cols_label(
             country = "Country",
             measles_total = "Measles Cases",
             rubella_total = "Rubella Cases") |>
  tab_style(
        style = cell_text(weight = "bold"),
        locations = list(cells_column_labels(), cells_row_groups())) |> 
  fmt_number(
        columns  = where(is.numeric),
        decimals = 1,
        use_seps = TRUE) |>
  data_color(
        columns = everything(),
        palette = "Oranges"
       )


```


## 2024 

```{r}
#| label: twenty four 

cases_year |>
  filter(year == 2024) |>
  select(country, measles_total, rubella_total) |>
  arrange(desc(measles_total)) |>
  slice_max(measles_total, n=15) |>
  gt() |>
  cols_label(
             country = "Country",
             measles_total = "Measles Cases",
             rubella_total = "Rubella Cases") |>
  tab_style(
        style = cell_text(weight = "bold"),
        locations = list(cells_column_labels(), cells_row_groups())) |> 
  fmt_number(
        columns  = where(is.numeric),
        decimals = 1,
        use_seps = TRUE) |>
  data_color(
        columns = everything(),
        palette = "Oranges"
       )


```

## 2025 

```{r}
#| label: twenty five 

cases_year |>
  filter(year == 2025) |>
  select(country, measles_total, rubella_total) |>
  arrange(desc(measles_total)) |>
  slice_max(measles_total, n=15) |>
  gt() |>
  cols_label(
             country = "Country",
             measles_total = "Measles Cases",
             rubella_total = "Rubella Cases") |>
  tab_style(
        style = cell_text(weight = "bold"),
        locations = list(cells_column_labels(), cells_row_groups())) |> 
  fmt_number(
        columns  = where(is.numeric),
        decimals = 1,
        use_seps = TRUE) |>
  data_color(
        columns = everything(),
        palette = "Oranges"
       )


```



::: 


# Yemen 



## Descriptions {width="50%"}

### Country Choice {height="50%"}
**Why Yemen?**    
Yemen is an **outlier** country with unusually high measles rates. In conjunction, their socio-economic history is complex, with synergistic conditions across access to healthcare, GDP and war. Over the last three years, they have ranked in the top 5 countries with the highest measles rates, while falling in the bottom 5 countries per GDP. Additionally, they have had a [civil war since 2014](https://www.britannica.com/event/Yemeni-Civil-War). Through this analysis, we aim to investigate the relationship between measles rates, GDP and civil war.

**Additional Data Sources**    

- [Worldometer](https://www.worldometers.info/gdp/yemen-gdp/) provides yearly GDP data from the World Bank and International Monetary Fund. The data is collected by WB and IMF through yearly reports and Worldometer contains GDP data for all of the world. The temporal availability of GDP depends on each country. For Yemen, Worldometer has published GDP data from 1990-2026. GDP is a critical measure for socio-economic conditions and is essential for evaluating the relationship between measles and economic status.      

- [Yemen Data Project](https://yemendataproject.org/data/) provides monthly aerial bombing data from 3/2015 - 4/2022 including locations, casualties and the number of bombs dropped. The data has been collected by the Yemen Data Project and is collected through open sources and cross-referenced using a wide range of information. These include local and international news agencies and media reports; social media accounts, including Twitter, Facebook, YouTube, and other video footage, and WhatsApp; reports from international and national NGOs; official records from local authorities; and reports by international human rights groups. This data is critical to highlighting the struggles of the people of Yemen.


### Map {height="50%"}

```{r}
leaflet() |>
  addProviderTiles("Esri.WorldStreetMap") |>
  setView(lng = 47.5, lat = 15.5, zoom = 5) |>
  setMaxBounds(lng1 = 40, lat1 = 11, lng2 = 56, lat2 = 20)

```


## Plots {width="50%"}

```{r}
yemen <- meta_data |>
  filter(country == "Yemen") |>
  select(country, year, month, measles_total, rubella_total)

#scrape table for yearly GDP data
url <- "https://www.worldometers.info/gdp/yemen-gdp/"

result <- tryCatch({
       read_html(url)
     }, error = function(e) {
       message("Failed for ", country, e$message)
       NULL
     })

gdp_data <- result |>
  html_elements("table") |>
  html_table() 

clean_me <- function(df) {
  df |>
    rename(GDP_USD = 2) |>
    select(Year, GDP_USD) |>
    mutate(GDP_USD = str_remove_all(GDP_USD, "\\$"),
    GDP_USD = str_remove_all(GDP_USD, ","),
    GDP_USD = as.numeric(GDP_USD))

}

gdp_wb <- clean_me(as.data.frame(gdp_data[1]))

gdp_imf <- clean_me(as.data.frame(gdp_data[3])) |>
  filter(Year < 2014)

# final gdp data to the 90s
yemen_gdp <- bind_rows(gdp_wb, gdp_imf)

# bombing data 2015 - 2022 
bomb <- read.csv('yemen-bombing.csv') |>
  mutate(Date = dmy(Date),
         month = month(Date),
  year = year(Date)) |>
  group_by(month, year) |>
  summarize(bombing_events = n(),
            total_casualty = sum(Fatalities + Woman.fatalities, na.rm = TRUE))


#final monthly yemen df (only bomb data 2015-2022)
yemen_monthly <- left_join(yemen, yemen_gdp, by = c("year" = "Year")) |>
  left_join(bomb, by = c("year", "month")) 
  

#yearly yemen measles calc
yemen_measles_yearly <- yemen |>
  group_by(year) |>
  summarize(measles_total = sum(measles_total),
            rubella_total = sum(rubella_total))

# final yearly yemen df
yemen_yearly <- bomb |>
  group_by(year) |>
  summarize(bombing_events = sum(bombing_events, na.rm=TRUE),
            total_casualty = sum(total_casualty, na.rm=TRUE)) |>
  right_join(yemen_measles_yearly, by = "year")


yemen_yearly |>
  ggplot(aes(x=year, y=measles_total, size = total_casualty)) +
  geom_point()

```



